Improving the portuguese hyphenation rules

### 2023-04-18 article/article.tex
updates from alines

\cite{rezende1987} created the first patters for Portuguese hyphenation in
\TeX{}. It was lastly updated in 2015 with the contributions from José João
Dias Almeida \cite{hyphpt}. The set has \NumberOfDefaultRules{}
rules\footnote{The set of rules for Portuguese hyphenation is short when
compared to rules in other languages. For example, English currently has 4938
rules, Russian has 7023 rules and German has 34011 rules.} and are able to
correctly hyphenate the majority of Portuguese words. We propose here to
analyse the performance of the default rules and add new rules to improve
hyphenation by correcting some issues and taking in account non-typical
patterns.

In order to apprise the performance of each set of rules, we need a list of
correctly hyphenated words that are used in a language. We use the frequency of
occurrence to avoid selecting really rare words. To build a large list with
such information, we relied on data collected from online dictionaries and
corpora of Portuguese. We also looked up Portuguese grammars to check on the
hyphenation (translineation) rules and we made some manual corrections on the
data.

In 1990, Portuguese went under a Language Orthographic Agreement. The adopting
transition period started in 2009 and ended as it turned mandatory in 2016.
Even after the agreement, some words still hold different spellings in the
participant countries. It is desirable to have hyphenation rules that might
work properly, despite those idiosyncrasies.  For example, the word
\emph{reception} might be written as \emph{receção} (Portugal) or
\emph{recepção} (Brazil); the word \emph{action} might be written as
\emph{acção} (Portugal) or \emph{ação} (Brazil); and the word \emph{project} might be
written as \emph{projecto} (Portugal) or \emph{projeto} (Brazil).

@@

We started using
\href{https://www.linguateca.pt/cetenfolha/index_info.html}{CETENFolha} as the
source corpus. Using only this corpus led to two problems: many words were
still missing and it was built on text from 1994, prior to the Orthographic
Agreement. We decided to incorporate the list of words from
\href{https://www.palavras.net/}{Palavras NET} and afterwards we also included
words from the \href{https://pt.wikipedia.org}{Portuguese Wikipedia} dump.
Although the list of words created was huge, we kept only those words we could
find their hyphenation on at least one of the following online dictionaries: 
\href{https://michaelis.uol.com.br/}{Michaelis},
\href{https://dicionario.priberam.org/}{Priberam},
\href{https://pt.wiktionary.org}{Wikcionário},
\href{https://aulete.com.br/}{Aulete},
\href{http://www.portaldalinguaportuguesa.org/}{Portal da Língua Portuguesa} e
\href{https://www.dicio.com.br/}{Dicio}.
This leads to the creation of a dictionary with \DictionarySize{} words.

Table \ref{tblstats} presents the number of words with a given number of
hyphenation found the dictionaries. It also presents the number of words that
have the highest agreed hyphenation. We will present some examples below. To
reference the cells in the table we will not consider the header line and
column. For example, the word \emph{como} was hyphenated by the six
dictionaries as \emph{co-mo}. It will add one to the counts on both lines
where the hyphenations number is 6 (first column). The word \emph{sua} was
hyphenated as \emph{su-a} in 4 dictionaries and a \emph{sua} in one
dictionary. It will add one in the first line, second column (since hyphenations
were found in 5 dictionaries). It will also add one on the second line, third
column (since the most frequent hyphenation was found in 4 dictionaries). It is
clear now why we might have the number of same hyphenations (specially for
small values) greater the number of hyphenations found.

@@

@@

%When there is no space left in line for a whole word, it might be split in two
%using a hyphen as an indication of such procedure. 
In Portuguese, hyphenations are allowed on syllables boundaries. According to
the Grammar \cite{cunha2016,bergstrom2011}, some rules might still apply:
\begin{enumerate}
    \item\label{rule-di-triphthong} diphthong or triphthong should not be split
	(e.g. \emph{m\underline{\emph{ui}}-to}, \emph{Pa-ra-g\underline{\emph{uai}}});
    \item\label{rule-unstressed} the sequences \emph{ia}, \emph{ie}, \emph{io}, \emph{oa}, \emph{ua},
	\emph{ue} and \emph{uo}, when in final unstressed position, should not be split 
	(e.g. \emph{gló-r\underline{\emph{ia}}}, \emph{vi-tó-r\underline{\emph{ia}}}, 
	\emph{cá-r\underline{\emph{ie}}}, \emph{es-pé-c\underline{\emph{ie}}}, 
	\emph{Má-r\underline{\emph{io}}}, \emph{má-g\underline{\emph{oa}}}, 
	\emph{ré-g\underline{\emph{ua}}}, \emph{tê-n\underline{\emph{ue}}}, 
	\emph{con-tí-g\underline{\emph{uo}}}, \emph{am-bí-g\underline{\emph{uo}}});
    \item\label{rule-c-clusters} consonant clusters starting a syllable should not be
	split (e.g. \emph{\underline{\emph{pn}}eu-má-ti-co}, \emph{\underline{\emph{ps}}i-có-lo-go}, 
	\emph{\underline{\emph{mn}}e-mô-ni-co});
    \item\label{rules-digraphs-ns} the digraphs \emph{ch}, \emph{lh}, \emph{nh}, \emph{gu}, \emph{qu} 
	should not be split (e.g. \emph{ra-\underline{\emph{ch}}ar},
	\emph{a-bro-\underline{\emph{lh}}os}, \emph{ma-\underline{\emph{nh}}ã},
	\emph{\underline{\emph{gu}}i-tar-ra}, \emph{\underline{\emph{qu}}in-to});
    \item\label{rule-nasalization} a vowel and its following nasalization marker (a graphic
	nasal consonant) should not be split (e.g. \emph{\underline{\emph{am}}-bição},
	\emph{m\underline{\emph{an}}-cha});
    \item\label{rule-rising} rising diphthong should not be split (e.g.
	\emph{a-b\underline{\emph{ai}}-xo}; \emph{c\underline{\emph{au}}-te-la}, 
	\emph{pa-p\underline{\emph{éi}}s}, \emph{cha-p\underline{\emph{éu}}}, 
	\emph{pre-f\underline{\emph{ei}}-to}, \emph{r\underline{\emph{eu}}-nir},
	\emph{n\underline{\emph{oi}}-te}; \emph{ca-la-b\underline{\emph{ou}}-ço}, 
	\emph{as-te-r\underline{\emph{ói}}-de}; 
	\emph{re-tri-b\underline{\emph{ui}}});

@@

>>>>

\cite{rezende1987} created the first patters for Portuguese hyphenation in
\TeX{}. It was lastly updated in 2015 with the contributions from José João
Dias Almeida \cite{hyphpt}. The set has \NumberOfDefaultRules{}
rules\footnote{The set of rules for Portuguese hyphenation is short when
compared to rules in other languages. For example, English currently has 4938
rules, Russian has 7023 rules and German has 34011 rules.} and are able to
correctly hyphenate the majority of Portuguese words. 
% - detalhar as regras de Rezende e Dias 
% - contabilizar a quantidade de palavras que não são contempladas pela proposta do autor
% - analisar a necessidade de excluir alguma regra (ao invés de apenas acrescentar novas)

In front of these cases, we propose here to analyse the performance of the default rules in order to add new rules to improve
hyphenation by correcting some issues and taking in account non-typical
patterns. We describe the methodology used in the section \ref{sec-methodology}.

\section{Methodology}\label{sec-methodology}

In order to apprise the performance of each set of rules, we need a list of
correctly hyphenated words that are used in Portuguese. For that, we use the
frequency of occurrence to avoid selecting really rare words. To build a large
list with such information, we relied on data collected from online
dictionaries and corpora of Portuguese. We also looked up Portuguese grammars
to check on the hyphenation (\emph{translineation}) rules and we made some manual
corrections on the data.

\subsection{Data collection}\label{sec-data-coll}

Initially we selected \href{https://www.linguateca.pt/cetenfolha/index_info.html}{CETENFolha} 
as the source corpus. Using only this corpus led to two problems: many words
were still missing and it was built on text from 1994, prior to the
Orthographic Agreement. This problem is a result of the Orthographic Agreement
of the Portuguese Language, carried out in 1990.

The Agreement's adopting transition period started in 2009 and ended as it
turned mandatory in 2016.  Even after the agreement, some words still hold
different spellings in the participant countries. It is desirable to have
hyphenation rules that might work properly, despite those idiosyncrasies.  For
example, the word \emph{reception} might be written as \emph{receção}
(Portugal) or \emph{recepção} (Brazil); the word \emph{action} might be written
as \emph{acção} (Portugal) or \emph{ação} (Brazil); and the word \emph{project}
might be written as \emph{projecto} (Portugal) or \emph{projeto} (Brazil).

@@

Because of these idiosyncrasies, we decided to incorporate the list of words from
\href{https://www.palavras.net/}{Palavras NET} and afterwards we also included
words from the \href{https://pt.wikipedia.org}{Portuguese Wikipedia} dump.
Although the list of words created was huge, we kept only those words we could
find their hyphenation on at least one of the following online dictionaries: 
\href{https://michaelis.uol.com.br/}{Michaelis},
\href{https://dicionario.priberam.org/}{Priberam},
\href{https://pt.wiktionary.org}{Wikcionário},
\href{https://aulete.com.br/}{Aulete},
\href{http://www.portaldalinguaportuguesa.org/}{Portal da Língua Portuguesa} e
\href{https://www.dicio.com.br/}{Dicio}.
This leads to the creation of a dictionary with \DictionarySize{} words.

Table \ref{tblstats} presents the number of words with a given number of
hyphenation found in the dictionaries. It also presents the number of words
that have the highest agreed hyphenation. For example, the word \emph{como}
was hyphenated by the six dictionaries as \emph{co-mo}. It will add one to the
counts on both lines where the hyphenations number is 6 (first column). The
word \emph{sua} was hyphenated as \emph{su-a} in 4 dictionaries, \emph{sua} in
one dictionary and one dictionary did not provide its hyphenation. Therefore,
it will add one in the first line, second column (since hyphenations were found
in 5 dictionaries). It will also add one on the second line, third column
(since the most frequent hyphenation was found in 4 dictionaries). It is clear
now why we might have the number of same hyphenations (specially for small
values) greater the number of hyphenations found.

@@

In summary, data from \emph{CETENFolha} and \emph{Palavras NET} were compiled.
Their words were hyphenated based on the hyphenation orignal \TeX{} rules and
contrasted with the dictionary results. The performance was appraised and the
errors systematized, in order to find in the literature possible rules that
could complement and improve it.

@@

%When there is no space left in line for a whole word, it might be split in two
%using a hyphen as an indication of such procedure. 
In Portuguese, hyphenations are allowed on syllables boundaries and, in general, follow phonological principles. 
According to
the Grammar \cite{cunha2016,bergstrom2011,cegalla2008}, some rules might still apply:
\begin{enumerate}
    \item\label{rule-di-triphthong} diphthong or triphthong should not be split
	(e.g. \emph{m\underline{\emph{ui}}-to}, \emph{Pa-ra-g\underline{\emph{uai}}});
    \item\label{rule-unstressed} the sequences \emph{ia}, \emph{ie}, \emph{io}, \emph{oa}, \emph{ua},
	\emph{ue} and \emph{uo}, when in final unstressed position, should not be split 
	(e.g. \emph{gló-r\underline{\emph{ia}}}, \emph{vi-tó-r\underline{\emph{ia}}}, 
	\emph{cá-r\underline{\emph{ie}}}, \emph{es-pé-c\underline{\emph{ie}}}, 
	\emph{Má-r\underline{\emph{io}}}, \emph{má-g\underline{\emph{oa}}}, 
	\emph{ré-g\underline{\emph{ua}}}, \emph{tê-n\underline{\emph{ue}}}, 
	\emph{con-tí-g\underline{\emph{uo}}}, \emph{am-bí-g\underline{\emph{uo}}});
    \item\label{rule-c-clusters} consonant clusters starting a syllable should not be
	split (e.g. \emph{\underline{\emph{pn}}eu-má-ti-co}, \emph{\underline{\emph{ps}}i-có-lo-go}, 
	\emph{\underline{\emph{mn}}e-mô-ni-co});
    \item\label{rules-digraphs-ns} the digraphs\footnote{One should not confuse consonant clusters 
	with digraphs, since the latter represent only one sound, while the former represent two or more sounds.}
	\emph{ch}, \emph{lh}, \emph{nh}, \emph{gu}, \emph{qu} 
	should not be split (e.g. \emph{ra-\underline{\emph{ch}}ar},
	\emph{a-bro-\underline{\emph{lh}}os}, \emph{ma-\underline{\emph{nh}}ã},
	\emph{\underline{\emph{gu}}i-tar-ra}, \emph{\underline{\emph{qu}}in-to});
    \item\label{rule-guqu} bigrams like \emph{gu} and \emph{qu} whose vowel \emph{u} is not 
	pronounced are never separated from the vowel or diphthong that follows it 
	(e.g. \emph{ne-\underline{\emph{gu}e}}, \emph{pe-\underline{\emph{qu}e}});
    \item\label{rule-nasalization} since they are digraphs, a vowel and its following 
	nasalization marker (a graphic nasal consonant) should not be split (e.g. \emph{\underline{\emph{am}}-bição},
	\emph{m\underline{\emph{an}}-cha});
    \item\label{rule-decreasing} decreasing diphthongs should not be split (e.g. 
	\emph{\underline{\emph{ai}}-ro-so}, \emph{ca-d\underline{\emph{ei}}-ra}, 
	\emph{o-ra-ç\underline{\emph{ão}}});
    \item\label{rule-rising} rising diphthong should not be split (e.g.
	\emph{a-b\underline{\emph{ai}}-xo}; \emph{c\underline{\emph{au}}-te-la}, 
	\emph{pa-p\underline{\emph{éi}}s}, \emph{cha-p\underline{\emph{éu}}}, 
	\emph{pre-f\underline{\emph{ei}}-to}, \emph{r\underline{\emph{eu}}-nir},
	\emph{n\underline{\emph{oi}}-te}; \emph{ca-la-b\underline{\emph{ou}}-ço}, 
	\emph{as-te-r\underline{\emph{ói}}-de}; 
	\emph{re-tri-b\underline{\emph{ui}}});
    \item\label{rule-singlev} disyllables whose syllable has a single vowel should 
	not be split (e.g. \emph{\underline{\emph{a}}to}, \emph{ru\underline{\emph{a}}}, 
	\emph{\underline{\emph{ó}}dio}, \emph{\underline{\emph{u}}nha});
    \item\label{rule-orphan} words with more than two syllables, when divided, cannot 
	isolate a syllable composed of a single vowel (e.g. \emph{\underline{\emph{a}}gos-to}, 
	\emph{la-go\underline{\emph{a}}}, \emph{\underline{\emph{i}}da-de});

@@

Systematizing the rules that guide syllable boundarie and, consequently,
hyphenation in Portuguese proved to be fundamental for the improvement of the
\TeX{} hyphenator rules, since they indicate the possible or prohibited contexts
of hyphenation. Expomos, a seguir, uma análise comparativa...
